\documentclass[addpoints,12pt,A4]{exam}

\printanswers

\title{Continuous Optimization: Problem Set 2}

\input{ExSetup}


\begin{document}
\maketitle

\begin{questions}

 \question
 Recall the definitions of $\alpha$-strong convexity and $\beta$-smoothness: for all $x,y$
 \[ \frac{\alpha}{2} \|y - x\|_{2}^{2} \leq f(y) - f(x) - \langle \nabla f(x), y - x \rangle \leq \frac{\beta}{2} \|y - x\|_{2}^{2} .   \]
 We will examine equivalent $1$-st and $2$-nd order definitions of these properties. For this question, assume $f$ is twice continuously differentiable. 
 \begin{enumerate}
    \item Show if $f : \R^{n} \to \R$ is $\alpha$-strongly convex or $\beta$-smooth, then the restriction $g(t) := f(x + t v)$ is $\alpha \|v\|_{2}^{2}$-strongly convex and $\beta \|v\|_{2}^{2}$-smooth for any $x, v \in \R^{n}$;
     \item Show $\alpha$-strong convexity and $\beta$-smoothness are equivalent to the conditions $\alpha I \preceq \nabla^{2} f(x)$ and $\nabla^{2} f(x) \preceq \beta I$, respectively. (Here the notation $A \succeq B$ means $A - B \succeq 0$ or equivalently $\forall v: \langle v, A v \rangle \geq \langle v, B v \rangle$)
     \item Show for an $\alpha$-strongly convex function, we have distance bound on the optimizer
     \[ \|x - x^{*}\|_{2} \lesssim \frac{\|\nabla f(x)\|_{2}}{\alpha} .  \]     
     \item Show $\beta$-smoothness is equivalent to Lipschitz-ness of the \emph{gradient}:
     \[ \forall x,y : \quad \|\nabla f(y) - \nabla f(x)\|_{2} \leq \beta \|y - x\|_{2} .   \]
     (You can use the following fact without proof: $0 \preceq H \preceq \beta I \iff \forall x: \|H x\|_{2} \leq \beta \|x\|_{2} \iff \forall x,y: |\langle x, H y \rangle| \leq \beta \|x\|_{2} \|y\|_{2}$)
 \end{enumerate}
 \begin{solution}
\end{solution}

 \question
 In the analysis of gradient descent for strongly convex and smooth functions we used the following crucial claim, known as the Polyak-Lojaseiwicz inequality (PL):
 \[ \|\nabla f(x)\|_{2}^{2} \gtrsim \alpha (f(x) - f(x^{*})) ,    \]
 where $x^{*}$ is the optimizer. 
     \begin{enumerate}
     \item  Show that $\alpha$-strong convexity implies this $\alpha$-PL condition. 
     \item Show that the analysis goes through if we assume only the PL condition as above, instead of strong convexity. 
     \item Find a function that satisfies $\alpha=1$ PL condition but is not $\alpha=1$-strongly convex. (Hint: consider convex quadratics.) 
     \item \textbf{Remark}: this condition even allows rigorous fast convergence analyses for certain non-convex functions!
     % \item (Difficult): Consider gradient flow: $\partial_{t} x_{t} := - \nabla f(x_{t})$. Show the PL-condition implies exponential convergence of this process, even without smoothness.  
 \end{enumerate}
 \begin{solution}
 \end{solution}


\question  (Q4 Ex4 from Daniel Dadush Course 2022)
Assume $\MA \MA \Tr$ is invertible. Give an explicit formula for $\nu^* := {\rm max} ~g(\nu) := \nu \Tr \vb - \|\MA \Tr \nu\|_2^2/2$ in terms of $\MA \MA \Tr$ and $\vb$. Prove that $\MA \Tr \nu^*$ is the optimal solution to the primal least-squares problem $\min \|\vx\|_2^2/2, \MA \vx = \vb$.  
\begin{solution}
\end{solution}

 \question
 Recall our characterization of quadratic convex functions from last homework:
 \[ \langle x, H x \rangle + \langle b, x \rangle + c  \]
 is convex iff $H \succeq 0$, i.e. $H$ is symmetric and $\forall v \in \R^{n}: \langle v, H v \rangle \geq 0$ (strictly convex iff $H \succ 0$, i.e. the inequality is strict for $v \neq 0$). In this question we will consider partitioned matrices
 \[ q(x,y) := \bigg\langle \begin{pmatrix} x \\ y \end{pmatrix},  \begin{pmatrix}
     A & B \\ B^{T} & C
 \end{pmatrix} \begin{pmatrix} x \\ y \end{pmatrix} \bigg\rangle ,  \]
 where $A, C$ are symmetric. 
 \begin{enumerate}
     \item Consider $f : \R^{n} \times \R^{m} \to \R$ be a convex function of variables $x \in \R^{n}, y \in \R^{m}$. Show $g(x) := \inf_{y} f(x,y)$ is also convex. 
     \item Let $q(x,y)$ be the quadratic defined above and assume $C \succ 0$. Show 
     \[ \inf_{y} q(x,y) = \langle x, (A - B C^{-1} B^{T}) x \rangle .  \]
     (Hint: use first order optimality conditions for $y$). 
     \item Show $q$ is strictly convex iff 
     \[ H := \begin{pmatrix}
     A & B \\ B^{T} & C
 \end{pmatrix} \succ 0 \]
 iff $C \succ 0$ and $A - B C^{-1} B^{T}$. 
 \item (Not a question, just a note): This last expression $A - B C^{-1} B^{T}$ is known as the \emph{Schur complement} of $H$ onto the $x$ variables. It can be seen as the result of Gaussian elimination of the $y$ variables and is extremely useful throughout math and computer science. 
 \end{enumerate}
 \begin{solution}
 \end{solution}
 

\question
We use the following lemma: for convex $\ell$ and convex differentiable $r$, let $f := \ell + r$ with $x_{*} = \arg\min f$. Then 
\[ f(x) = \ell(x) + r(x) \geq f(x_{*}) + D_{r}(x,x_{*}) = \ell(x_{*}) + r(x_{*}) + D_{r}(x,x_{*}) .    \]
\begin{enumerate}
\item Prove the Bregman Pythagorean result: for strictly convex $r : \R^{n} \to \R$, let $\Pi_{K}(x) := \arg\min_{z \in K} D_{r}(z,x)$, then 
\[ \forall z \in K: \quad D_{r}(z, x) \geq D_{r}(\Pi(x), x) + D_{r}(z, \Pi(x)) .    \]
\item Let $f(z) := D_{r}(z,x)$, then $D_{f} = D_{r}$, i.e. 
\[ D_{f}(z,y) = D_{r}(z,x) - D_{r}(y,x) - \langle \nabla r(y) - \nabla r(x), z - y \rangle = D_{r}(z,y) .   \]
% \item Duality: $D_{\phi}(y,x) = D_{\phi^{*}}(\nabla \phi(x), \nabla \phi(y))$
\end{enumerate} 

\begin{solution}
\end{solution}



\question (Lap Chi Lau Course CS798, HW2 Q1)
We analyze the multiplicative weights method for solving the following linear program:
\[ \min \sum_{i \in [n]} x_{i} \qquad \text{s.t.} \qquad A x \geq \vec{1}, \quad x \geq 0 ,   \]
where $A \in \{0,1\}^{m \times n}$, i.e. every entry of $A$ is either $0$ or $1$. Design an algorithm to solving this linear program using the multiplicative weights update method (including the design of the oracle), and analyze its total running time.
\begin{solution}
    
\end{solution}


  \question
  In our proof of the lower bound for first-order methods for convex Lipschitz functions we made the following technical assumption: the initial point $x_{0} = 0$ and the queries $x_{t} \in \text{span}\{x_{0}, g_{0}, ..., g_{t-1}\}$. In this question we show how to design an \emph{adaptive} adversarial gradient oracle to remove this assumption. 
  \begin{enumerate}
      \item Show that we can remove the assumption $x_{0} = 0$ by shifting the function based on the initial query. 
      \item Show that for each query sequence $x_{0}, ... x_{T}$ with $T < n$ we can design a function of the form
      \[ f_{T}(x) = \max_{i \in [n]} \langle g_{i}, x - x_{0} \rangle  \]
      such that $f(x_{t}) \leq 0$ for all $t \in [T]$. 
      \item Show that we can choose functions $\{f_{0}, ..., f_{T}\}$ as above in a \emph{consistent} way: 
      \[ \forall s \leq t < n: f_{t}(x_{s}) = f_{s}(x_{s}) . \] Conclude that we can remove the technical assumption about queries. 
  \end{enumerate}
 \begin{solution}
 \end{solution}



\question (Lap Chi Lau Course CS798, HW2 Q1)
Suppose you are given an `accelerated' algorithm $\mathcal{A}$ for minimizing strongly convex and smooth functions: given $f$ that is $\alpha$-strongly convex and $\beta$-smooth, the algorithm outputs $x_{\text{alg}}$ satisfying $f(x_{\text{alg}}) - f(x^{*}) \leq \eps$ in 
\[ T \lesssim \sqrt{ \frac{\beta}{\alpha}} \cdot  \log \bigg( \frac{f(x_{0}) - f(x^{*})}{\eps} \bigg) \]
iteration, where $x^{*} = \arg\min_{x} f(x)$ (unknown), and $x_{0}$ is the initial point. 

\begin{itemize}
    \item You would like to use algorithm $\mathcal{A}$, but apply it to $f$ that is \emph{not strongly convex}. Prove that you can still use $\mathcal{A}$ as a black box to give an `accelerated' algorithm with the following guarantee: for convex $f$ that is $\beta$-smooth and minimizer $x^{*} = \arg\min_{x} f(x)$, find output $x$ such that $f(x) - f(x^{*}) \leq \eps$ in 
    \[ T \lesssim \sqrt{\frac{\beta \|x_{0} - x^{*}\|_{2}^{2}}{\eps}} \cdot \log \bigg( \frac{f(x_{0}) - f(x^{*})}{\eps} \bigg)   \]
    iterations, where again $x_{0}$ is the initial point. 
    You are allowed to assume that you know the values $\|x_{0} - x^{*}\|_{2}$ and $f(x_{0}) - f(x^{*})$. 

    \item Use this to derive the lower-bound for first-order methods for smooth convex functions via the lower-bound for smooth and strongly convex functions. 
\end{itemize}
\begin{solution}
    
\end{solution}



\end{questions}
\end{document}
